Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-01-28 doc/invt.tex
editing invt.tex

We first approximate Eq. \eqref{eq:wl_update}
by a differential equation
%
\begin{equation}
  \dot v_i(t)
  =
  h_i(t) \frac{ \alpha(t) } { p_i },
  \label{eq:vt_diffeq}
\end{equation}
%
where
$\dot v_i(t) \equiv dv_i(t)/dt$,
%
and $h_i(t) = \delta_{i, i(t)}$
is the instantaneous histogram,
which is equal to $1.0$
for the current bin $i(t)$
or zero otherwise.

@@

\subsubsection{One-variable problem}

@@

To do so, we first recall
the formal solution of Eq. \eqref{eq:dxdt_alpha}:
%
\begin{equation}
x(t) = x(0) \, e^{-\lambda \, q(t)}
+ \int_0^t \dot u\bigl( q(t') \bigr) \, \xi(t') \, dt',
\label{eq:xt_solution}
\end{equation}
%
where,
%
\begin{equation}
q(t) \equiv \int_0^t \alpha(t') \, dt',
\label{eq:qt_definition}
\end{equation}
%
and
%
\begin{align}
u(z)
&\equiv
e^{-\lambda \, q(t) + \lambda \, z}.
\label{eq:ut_definition}
\end{align}
So
\begin{align}
\left\langle x^2(t) \right\rangle
&=
\int_0^t \int_0^t
  \dot u\bigl( q(t') \bigr) \,
  \dot u\bigl( q(t'') \bigr) \,
  \kappa(t' - t'') \, dt'' \, dt'
\notag
\\
&\hphantom{=}
+\langle x^2(0) \rangle e^{-2 \, \lambda \, q(t)}.
\label{eq:x2t_average}
\end{align}

@@

We now consider the multiple-bin scheme,
Eq. \eqref{eq:mbin_update}.
%
In this case,
upon a visit to bin $j$,
we update not only the bias potential there,
but also those for a few neighboring bins $i$.
%
The update magnitudes are given by a matrix $\mathbf w$
with elements $w_{ij}$.
%
Then, Eq. \eqref{eq:vt_diffeq} is generalized to
\begin{equation}
  \dot v_i(t) =
  \sum_{j=1}^n \alpha(t) \, h_j(t) \frac{ w_{ij} } { p_j }.
  \label{eq:vt_diffeq_mbin}
\end{equation}

@@

\subsubsection{Linearization and decoupling}

@@

\note{The proof of Eq. \eqref{eq:dvbardt},
$$
\begin{aligned}
  \dot{ \bar v }
  &= \sum_{i = 1}^n p_i \, \dot v_i
  \stackrel{\mathrm{Eq.\; \eqref{eq:vt_diffeq_mbin}}}
  {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \sum_{j = 1}^n \alpha(t) \, \frac{ h_j(t) } { p_j } \sum_{ i = 1 }^{n} p_i \, w_{ij}
  \\
  &
  \stackrel{\mathrm{Eq.\; \eqref{eq:w_balance}}}
  {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \sum_{j = 1}^n \alpha(t) \, h_j(t) = \alpha(t).
\end{aligned}
$$
Equation \eqref{eq:px_sum} follows directly from the definition
Eq. \eqref{eq:x_def}.
}

Now by diagonalizing the matrix $\mathbf w$
using the eigenvectors defined in Eqs. \eqref{eq:eig_w},
\eqref{eq:eig_orthonormal_cols},
and
\eqref{eq:eig_orthonormal_rows},
we get a set of decoupled equations
%
\begin{equation}
\dot y_i(t)
=
-\alpha(t) \, \lambda_i
[y_i(t) - \eta_i(t)],
\label{eq:yt_diffeq}
\end{equation}
%
where
\begin{align}
  y_i &= \sum_{j=1}^n \phi_{ij} \, x_j,
  \label{eq:y_def}
  \\
  \eta_i &= \sum_{j=1}^n \phi_{ij} \frac{ \zeta_j}{ p_j}.
  \label{eq:eta_def}
\end{align}

\subsubsection{Minimization of the error function}

The error function can be obtained
from Eq. \eqref{eq:eig_orthonormal_cols}
as
\begin{align}
  \Err
  &\equiv
  \sum_{i = 1}^n p_i \, x_i^2
  =
  \sum_{i, j, k=1}^n \phi_{ki} \, \phi_{kj} \, x_i \, x_j
  =
  \sum_{k = 1}^n y_k^2.
  \label{eq:y2_sum}
\end{align}
%
Note that there is always one mode, say
$y_1 = \sum_{j=1}^n p_j \, x_j$
(corresponding to $\lambda_1 = 1$),
vanishes identically by Eq. \eqref{eq:px_sum}.
%
Thus, we can start the sum in Eq. \eqref{eq:y2_sum}
from $k = 2$.

In analogous to Eq. \eqref{eq:x2t_average},
we can rewrite Eq. \eqref{eq:y2_sum} as
%
\begin{align}
  \Err
  %&= \sum_{i = 2}^n \langle y_i^2 \rangle
  %\notag
  %\\
  &=
  \int_0^t \int_0^t
  \sum_{i = 2}^n
  \dot u_i\bigl( q(t') \bigr) \,
  \dot u_i\bigl( q(t'') \bigr) \,
  \kappa_i(t' - t'') \, dt' \, dt'',
  \label{eq:error_mbin}
\end{align}
%
where
\begin{align*}
  u_i(z)
  &\equiv
  e^{\lambda_i \, [z - q(t)]},
  \\
  \kappa_i(t - t')
  &\equiv
  \left\langle
    \eta_i(t) \, \eta_i(t')
  \right\rangle.
\end{align*}
%
\note{The extremal condition of $\Err$ is given by
%
\begin{align}
\sum_{i=2}^n
\lambda_i \, u_i\bigl( q(\tau) \bigr)
\int_0^t
\ddot u_i\bigl( q(t') \bigr) \, \kappa_i(t' - \tau) \, dt' = 0.
\label{eq:optimal_mbin}
\end{align}
%
If all $\lambda_i$, hence all $u_i$, are the same,
the above condition can be satisfied
by Eq. \eqref{eq:ddu_eq_0},
which is the single-bin case.
%
Unfortunately, the above expression
appears to be useless in a general setting.
}

@@

In the asymptotic regime,
we can approximate $\kappa_i(t)$ as
%
\begin{equation}
  \kappa_i(t) \approx \Gamma_i \, \delta(t)
  \label{eq:kappa_delta}
\end{equation}
%
where,
\begin{equation}
  \Gamma_i = \int_{-\infty}^\infty \kappa_i(t) \, dt,
  \notag
  %\label{eq:Gamma_integral}
\end{equation}
or in discrete time,
\begin{equation}
  \Gamma_i = \sum_{t = -\infty}^\infty \kappa_i(t).
  \label{eq:Gamma_sum}
\end{equation}
%
Then, Eq. \eqref{eq:error_mbin} is simplified as
%
\begin{align}
  \Err
  &=
  \int_0^t
  \sum_{i = 2}^n
  \Gamma_i \, \dot u_i^2\bigl( q(t') \bigr) \, dt'.
  \label{eq:error_mbin1}
\end{align}
%
The next order correction to Eq. \eqref{eq:kappa_delta}
would lead to a correction to $\Err$
that is about $\alpha_{\max}^2$ times as large.

\note{This is equivalent to saying that the Fourier transform
  $$
  \tilde \kappa_i(\omega) = \Gamma_i
  $$
  is a roughly constant.
  %
  For the next order correction we would have
  $$
  \tilde \kappa_i(\omega) = \Gamma_i + \Gamma^{(2)}_i \omega^2 + \dots,
  $$
  which leads to a correction to $\Err$
  $$
  \Gamma^{(2)}_i
  \int_0^t \ddot u_i^2\bigl( q(t') \bigr) \, dt'
  \propto
  \alpha^2 \, \Err.
  $$
  This correction is of order $\alpha^2$
  compared to the main contribution in Eq. \eqref{eq:error_mbin1}.
  %
  Thus, we can assume Eq. \eqref{eq:kappa_delta},
  once $\alpha(t)$ has dropped below a certain level.
}

\note{
  The extremal property, Eq. \eqref{eq:optimal_mbin},
  now becomes
  %
  \begin{align}
    \sum_{i=1}^n
      \Gamma_i \, \lambda_i \,
      u_i\bigl( q(\tau) \bigr) \,
      \ddot u_i\bigl( q(\tau) \bigr) = 0.
    \label{eq:optimal_mbin1}
  \end{align}
  %
  If we interpret $u_i$ as the position,
  and $\Gamma_i \, \lambda_i$ as the mass,
  then the right-hand side gives the virial.
  %
  Thus, the extremal condition demands
  the total virial to be zero.
  $$\,$$
}

@@

We can evaluate the $\Gamma_i$'s
in two special case.
%
If the sampling is perfect,
Eq. \eqref{eq:kappa_delta} is exact, and
%
\begin{equation}
  \Gamma_i = 1.
  \label{eq:Gamma_perfect}
\end{equation}
%
In this case,
the bin index $i$ is an independent random variable
at each time step $t$, and
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \left( p_i \, \delta_{ij} - p_i \, p_j \right) \, \delta(t - t'),
  \label{eq:zz_perfect}
\end{equation}
%
where $\delta(t - t')$ is understood as an adaption of $\delta_{t, t'}$
from the discrete time case to the continuous limit. % hereinafter.
%
\note{To show Eq. \eqref{eq:zz_perfect}, we first note that
  with $t \ne t'$ (going back to the discrete time case),
  $$
  \begin{aligned}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  &=
  \langle \zeta_i(t)  \rangle
  \langle \zeta_j(t') \rangle
  \\
  &=
  \bigl\langle h_i(t)  - \langle h_i(t)  \rangle \bigr\rangle
  \bigl\langle h_j(t') - \langle h_j(t') \rangle \bigr\rangle
  =
  0.
  \end{aligned}
  $$
  Next, for $t = t'$, we have
  $$
  \zeta_i(t)
  = h_i(t) - \langle h_i(t) \rangle
  = \delta_{i, i(t)} - p_i,
  $$
  and
  $$
  \begin{aligned}
  \left\langle
    \zeta_i(t) \, \zeta_j(t)
  \right\rangle
  &=
  \left\langle
    \delta_{i, i(t)} \, \delta_{j, i(t)}
  \right\rangle
  -
  p_i \, p_j
  \\
  &=
  \left\langle
    \delta_{i, i(t)}
  \right\rangle
  \delta_{i, j}
  -
  p_i \, p_j
  =
  p_i \, \delta_{ij}
  - p_i \, p_j.
  \end{aligned}
  $$
  %
  The average
  $\left\langle \delta_{i, i(t)} \right\rangle$ is $p_i$,
  because as $i(t)$ goes through $1$ to $n$,
  $\delta_{i, i(t)}$ gives $1$
  if $i(t)$ matches $i$ with probability $p_i$,
  or $0$ otherwise.
}
%
Thus, from Eqs. \eqref{eq:eig_orthonormal_rows} and \eqref{eq:eta_def},
we get, for $i \ge 2$,
%
\begin{equation}
  \kappa_i(t - t')
  =
  \sum_{j,k = 1}^n
  \frac{ \phi_{ij} } { p_j }
  \frac{ \phi_{ik} } { p_k }
  \langle \zeta_j(t) \, \zeta_k(t') \rangle
  %
  = \delta(t - t'),
  \label{eq:kappa_perfect}
\end{equation}
%
\note{To show Eq. \eqref{eq:kappa_perfect},
  $$
  \begin{aligned}
  \kappa_i(t - t')
  &=
  \sum_{j,k = 1}^n
  \frac{ \phi_{ij} } { p_j }
  \frac{ \phi_{ik} } { p_k }
  \langle \zeta_j(t) \, \zeta_k(t') \rangle
  \\
  %
  &=
  \sum_{j,k = 1}^n
  \frac{ \phi_{ij} } { p_j }
  \frac{ \phi_{ik} } { p_k }
  \left(
    p_j \, \delta_{j, k} - p_j \, p_k
  \right)
  \, \delta(t - t')
  \\
  &=
  \left(
    \sum_{j = 1}^n
    \frac{ \phi_{ij}^2 } { p_j }
    -
    \sum_{j = 1}^n \phi_{ij}
    \sum_{k = 1}^n \phi_{ik}
  \right)
  \, \delta(t - t')
  \\
  &=
  (1 - 0 \cdot 0) \, \delta(t - t').
  \end{aligned}
  $$
  where,
  we have used
  Eq. \eqref{eq:eta_def}
  on the first line,
  %
  Eq. \eqref{eq:zz_perfect}
  on the second line,
  and
  Eq. \eqref{eq:eig_orthonormal_rows}
  on the last line.
  %
  Note for $i \ge 2$,
  $$
  \sum_{j = 1}^n \phi_{ij}
  =
  \sum_{j = 1}^n \frac{ \phi_{ij} \, \phi_{1j} } { p_j }
  =
  \delta_{i1}
  =
  0.
  $$
  as the mode $1$, or the first row, $\phi_{1j}$,
  is given by the equilibrium distribution $p_j$.
}

Another special case is the one-step process\cite{vankampen}:
in each step, the system hops to the previous bin $i - 1$
or the next bin $i + 1$ with equal probability $\frac 1 2$,
ignoring out-of-boundary transitions.
%
The transition matrix is
%
\begin{equation}
\arraycolsep=3.6pt\def\arraystretch{1.4}
\mathbf T
=
\left(
  \begin{array}{cccccccc}
    \frac 1 2 & \frac 1 2 & 0 & \dots & 0 \\
    \frac 1 2 & 0         & \frac 1 2 & \dots & 0 \\
    \vdots & &  & & \vdots \\
    0 & \dots & \frac 1 2 & 0  & \frac 1 2 \\
    0 & \dots & 0 & \frac 1 2 & \frac 1 2
  \end{array}
\right).
\label{eq:T_tridiag}
\end{equation}
%
We shall further assume that
the eigenvectors of $\mathbf w$
coincide with those of $\mathbf T$,
while the eigenvalues may differ
(this is true for a class of matrices
discussed in Sec. \ref{sec:band-matrix}).
%
The eigenvalues of $\mathbf T$ are
$$
\Lambda_i = 1 - 2 \sin^2 \frac{ (i - 1) \, \pi } { 2 \, n }.
$$
Note that these $\Lambda_i$'s are
not to be confused with the eigenvalues of $\mathbf w$,
$\lambda_i$'s.
%
The correlation function $\kappa_i(t)$
decays as $\kappa_i(0) \, \Lambda_i^t$,
with the same-time value $\kappa_i(0) = 1$ being the same as
that in perfect sampling.
%
Thus, from Eq. \eqref{eq:Gamma_sum}, we get
%
\begin{equation}
\Gamma_i
=
1 + 2 \, \sum_{t = 1}^\infty \Lambda_i^t
=
\frac{ 1 + \Lambda_i } { 1 - \Lambda_i }
=
\cot^2 \left[ \frac{ (i - 1) \, \pi } { 2 \, n } \right].
\label{eq:Gamma_onestep}
\end{equation}
%
We can use this case to model local sampling processes.

%
%We can study two limiting cases.
%
%For $\tau \ll t$,
%the dominant contribution comes from
%the term(s) with the smallest eigenvalue $\lambda_i = \lambda_n$,
%and
%$$
%\ddot u_n(\tau) = 0,
%\quad
%\dot u_n(\tau) = c,
%\quad
%\mathrm{and}\;
%u_n(\tau) = c \, (\tau + t_0),
%$$
%which means
%$$
%\alpha(\tau) \approx \frac{1}{\lambda_n \, (\tau + t_0)}.
%$$

@@

We can find an analytical expression
of the optimal value of $\lambda$
in the limit of large $n$.
%
In this case,
we convert Eq.
\eqref{eq:y2_sum_perfect_invt}
%and
%\eqref{eq:optimal_lambda_approx}
to an integral,
$\sum_i \to \frac{2 \, n}{\pi} \int dp$, and
$$
\begin{aligned}
\Err_2
&=
\frac{2 \, n}{\pi \, t}
\int_0^{\pi/2}
  \frac{    \Gamma(p) \, \left(1 - 4 \, \mu_1 \, \sin^2 p \right)^2    }
       { \lambda \, \left(2 - 8 \, \mu_1 \, \sin^2 p - \lambda \right) }
\, dp.
\end{aligned}
$$

@@

Another common updating scheme
in metadynamics is a Gaussian wave.
%
Below we show that this scheme
is stable, although
it might be less effective
in reducing local noises.

@@

We have shown that
without a priori information,
the single-bin scheme is asymptotically
the best among all updating scheme
in term of convergence.

>>>>

We first approximate Eq. \eqref{eq:wl_update}
by a differential equation
%
\begin{equation}
  \dot v_i(t)
  =
  \alpha(t) \, \frac{ h_i(t) } { p_i },
  \label{eq:vt_diffeq}
\end{equation}
%
where
$\dot v_i(t) \equiv dv_i(t)/dt$,
%
and $h_i(t) = \delta_{i, i(t)}$
is the instantaneous histogram,
which is equal to $1.0$
for the current bin $i(t)$
or zero otherwise.

@@

\subsubsection{Optimization of a one-variable problem}

@@

To do so, we first recall
the formal solution of Eq. \eqref{eq:dxdt_alpha}:
%
\begin{equation}
x(t) = x(0) \, e^{-\lambda \, q(t)}
+ \int_0^t \dot u\bigl( q(t') \bigr) \, \xi(t') \, dt',
\label{eq:xt_solution}
\end{equation}
%
where,
%
\begin{equation}
q(t) \equiv \int_0^t \alpha(t') \, dt',
\label{eq:qt_definition}
\end{equation}
%
and
%
\begin{align}
u(z)
&\equiv
e^{-\lambda \, q(t) + \lambda \, z}.
\label{eq:ut_definition}
\end{align}
So
\begin{align}
&\left\langle x^2(t) \right\rangle
=
\langle x^2(0) \rangle e^{-2 \, \lambda \, q(t)}
\notag
\\
&\qquad
+
\int_0^t \int_0^t
  \dot u\bigl( q(t') \bigr) \,
  \dot u\bigl( q(t'') \bigr) \,
  \kappa(t' - t'') \, dt'' \, dt'.
\label{eq:x2t_average}
\end{align}

@@

We now consider the multiple-bin scheme,
Eq. \eqref{eq:mbin_update}.
%
In this case,
upon a visit to bin $j$,
we update not only the bias potential there,
but also those for a few neighboring bins $i$.
%
The update magnitudes are given by a matrix $\mathbf w$
with elements $w_{ij}$.
%
Then, Eq. \eqref{eq:vt_diffeq} is generalized to
\begin{equation}
  \dot v_i(t)
  =
  \alpha(t) \sum_{j=1}^n \, w_{ij} \frac{ h_j(t) } { p_j }.
  \label{eq:vt_diffeq_mbin}
\end{equation}

@@

\subsubsection{Linearization and decoupling to eigenmodes}

@@

\note{The proof of Eq. \eqref{eq:dvbardt},
$$
\begin{aligned}
  \dot{ \bar v }
  &=
  \sum_{i = 1}^n p_i \, \dot v_i
  %
  \stackrel{\mathrm{Eq.\; \eqref{eq:vt_diffeq_mbin}}}
  {=\joinrel=\joinrel=\joinrel=\joinrel=}
  %
  \sum_{j = 1}^n \alpha(t) \, \frac{ h_j(t) } { p_j }
                 \sum_{ i = 1 }^{n} p_i \, w_{ij}
  \\
  &
  \stackrel{\mathrm{Eq.\; \eqref{eq:w_balance}}}
  {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \sum_{j = 1}^n \alpha(t) \, h_j(t)
  =
  \alpha(t).
\end{aligned}
$$
Equation \eqref{eq:px_sum} follows directly from the definition
Eq. \eqref{eq:x_def}.
}

Now by diagonalizing the matrix $\mathbf w$
using the eigenvectors defined in Eqs. \eqref{eq:eig_w},
\eqref{eq:eig_orthonormal_cols},
and
\eqref{eq:eig_orthonormal_rows},
we get a set of decoupled equations
%
\begin{equation}
\dot y_k(t)
=
-\alpha(t) \, \lambda_k \, [y_k(t) - \eta_k(t)],
\label{eq:yt_diffeq}
\end{equation}
%
where
\begin{align}
  y_k &= \sum_{j=1}^n \phi_{kj} \, x_j,
  \label{eq:y_def}
  \\
  \eta_k &= \sum_{j=1}^n \phi_{kj} \frac{ \zeta_j}{ p_j}.
  \label{eq:eta_def}
\end{align}

\subsubsection{Error function}

The error function can be obtained
from Eq. \eqref{eq:eig_orthonormal_cols}
as
\begin{align}
  \Err
  \equiv
  \sum_{i = 1}^n p_i \,
                 \langle x_i^2 \rangle
  =
  \sum_{i, j, k=1}^n \phi_{ki} \, \phi_{kj} \,
                     \langle x_i \, x_j \rangle
  =
  \sum_{k = 1}^n \langle y_k^2 \rangle,
  \label{eq:y2_sum}
\end{align}
%
which is a superposition of the components of the modes.
%
Note that there is always one mode, say
$y_1 = \sum_{j=1}^n p_j \, x_j$
(corresponding to $\lambda_1 = 1$),
vanishes identically by Eq. \eqref{eq:px_sum}.
%
Thus, we can start the sum in Eq. \eqref{eq:y2_sum}
from $k = 2$.

In analogous to Eq. \eqref{eq:x2t_average},
we have, from Eq. \eqref{eq:y2_sum},
%
\begin{equation}
  \Err
  =
  \Err_R + \Err_A,
  \label{eq:error_split}
\end{equation}
%
which is the sum of the residual error
%
\begin{equation}
  \Err_R
  =
  \sum_{k = 2}^n
    \left\langle y_k^2(0) \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(t) },
  \label{eq:error_res}
\end{equation}
%
from the decay of the initial errors,
and the asymptotic error
\begin{equation}
  \Err_A
  =
  \int_0^t \int_0^t
  \sum_{k = 2}^n
    \dot u_k\bigl( q(t') \bigr) \,
    \dot u_k\bigl( q(t'') \bigr) \,
    \kappa_k(t' - t'') \, dt' \, dt'',
  \label{eq:error_asym}
\end{equation}
%
where,
\begin{align*}
  u_k(z)
  &\equiv
  e^{\lambda_k \, [z - q(t)]},
  \\
  \kappa_k(t - t')
  &\equiv
  \left\langle
    \eta_k(t) \, \eta_k(t')
  \right\rangle.
\end{align*}
%
\note{The extremal condition of $\Err_A$ is given by
%
\begin{align}
\sum_{k = 2}^n
\lambda_k \, u_k\bigl( q(\tau) \bigr)
\int_0^t
\ddot u_k \bigl( q(t') \bigr) \, \kappa_k(t' - \tau) \, dt' = 0.
\label{eq:optimal_mbin}
\end{align}
%
If all $\lambda_k$, hence all $u_k$, are the same,
the above condition can be satisfied
by Eq. \eqref{eq:ddu_eq_0},
which is the single-bin case.
%
Unfortunately, the above expression
appears to be useless in a general setting.
}

@@

In the asymptotic regime,
we can approximate $\kappa_k(t)$ as
%
\begin{equation}
  \kappa_k(t) \approx \Gamma_k \, \delta(t)
  \label{eq:kappa_delta}
\end{equation}
%
where,
\begin{equation}
  \Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt,
  \notag
  %\label{eq:Gamma_integral}
\end{equation}
or in discrete time,
\begin{equation}
  \Gamma_k = \sum_{t = -\infty}^\infty \kappa_k(t).
  \label{eq:Gamma_sum}
\end{equation}
%
Then, Eq. \eqref{eq:error_asym} is simplified as
%
\begin{align}
  \Err_A
  &=
  \int_0^t
  \sum_{k = 2}^n
  \Gamma_k \, \dot u_k^2\bigl( q(t') \bigr) \, dt'.
  \label{eq:error_asym1}
\end{align}
%
The next order correction to Eq. \eqref{eq:kappa_delta}
would lead to a correction to $\Err$
that is about $\alpha_{\max}^2$ times as large.

\note{This is equivalent to saying that the Fourier transform
  $$
  \tilde \kappa_k(\omega) = \Gamma_k
  $$
  is a roughly constant.
  %
  For the next order correction we would have
  $$
  \tilde \kappa_k(\omega) = \Gamma_k + \Gamma^{(2)}_k \omega^2 + \dots,
  $$
  which leads to a correction to $\Err$
  $$
  \Gamma^{(2)}_k
  \int_0^t \ddot u_k^2\bigl( q(t') \bigr) \, dt'
  \propto
  \alpha^2 \, \Err.
  $$
  This correction is of order $\alpha^2$
  compared to the main contribution in Eq. \eqref{eq:error_asym1}.
  %
  Thus, we can assume Eq. \eqref{eq:kappa_delta},
  once $\alpha(t)$ has dropped below a certain level.
}

\note{
  The extremal property, Eq. \eqref{eq:optimal_mbin},
  now becomes
  %
  \begin{align}
    \sum_{k = 1}^n
      \Gamma_k \, \lambda_k \,
      u_k\bigl( q(\tau) \bigr) \,
      \ddot u_k\bigl( q(\tau) \bigr) = 0.
    \label{eq:optimal_mbin1}
  \end{align}
  %
  If we interpret $u_k$ as the position,
  and $\Gamma_k \, \lambda_k$ as the mass,
  then the right-hand side gives the virial.
  %
  Thus, the extremal condition demands
  the total virial to be zero.
  $$\,$$
}

@@

We can evaluate the $\Gamma_k$'s
in two special case.
%
If the sampling is perfect,
Eq. \eqref{eq:kappa_delta} is exact, and
%
\begin{equation}
  \Gamma_k = 1.
  \label{eq:Gamma_perfect}
\end{equation}
%
In this case,
the bin index $k$ is an independent random variable
at each time step $t$, and
%
\begin{equation}
  \left\langle
    \zeta_i(t) \, \zeta_j(t')
  \right\rangle
  =
  \left( p_i \, \delta_{ij} - p_i \, p_j \right) \, \delta(t - t'),
  \label{eq:zz_perfect}
\end{equation}
%
where $\delta(t - t')$ is understood as an adaption of $\delta_{t, t'}$
from the discrete time case to the continuous limit. % hereinafter.
%
\note{To show Eq. \eqref{eq:zz_perfect}, we first note that
  with $t \ne t'$ (going back to the discrete time case),
  $$
  \begin{aligned}
  \left\langle
    \zeta_i(t) \, \zeta_j(t')
  \right\rangle
  &=
  \left\langle
    \zeta_i(t)
  \right\rangle
  \left\langle
    \zeta_j(t')
  \right\rangle
  \\
  &=
  \bigl\langle h_i(t)  - \langle h_i(t)  \rangle \bigr\rangle
  \bigl\langle h_j(t') - \langle h_j(t') \rangle \bigr\rangle
  =
  0.
  \end{aligned}
  $$
  Next, for $t = t'$, we have
  $$
  \zeta_i(t)
  = h_i(t) - \langle h_i(t) \rangle
  = \delta_{i, i(t)} - p_i,
  $$
  and
  $$
  \begin{aligned}
  \left\langle
    \zeta_i(t) \, \zeta_j(t)
  \right\rangle
  &=
  \left\langle
    \delta_{i, i(t)} \, \delta_{j, i(t)}
  \right\rangle
  -
  p_i \, p_j
  \\
  &=
  \left\langle
    \delta_{i, i(t)}
  \right\rangle
  \delta_{i, j}
  -
  p_i \, p_j
  =
  p_i \, \delta_{ij}
  - p_i \, p_j.
  \end{aligned}
  $$
  %
  The average
  $\left\langle \delta_{i, i(t)} \right\rangle$ is $p_i$,
  because as $i(t)$ goes through $1$ to $n$,
  $\delta_{i, i(t)}$ gives $1$
  if $i(t)$ matches $i$ with probability $p_i$,
  or $0$ otherwise.
}
%
Thus, from Eqs. \eqref{eq:eig_orthonormal_rows} and \eqref{eq:eta_def},
we get, for $k \ge 2$,
%
\begin{equation}
  \kappa_k(t - t')
  =
  \sum_{j,l = 1}^n
  \frac{ \phi_{kj} } { p_j }
  \frac{ \phi_{kl} } { p_l }
  \langle \zeta_j(t) \, \zeta_l(t') \rangle
  %
  = \delta(t - t'),
  \label{eq:kappa_perfect}
\end{equation}
%
\note{To show Eq. \eqref{eq:kappa_perfect},
  $$
  \begin{aligned}
  \kappa_k(t - t')
  &=
  \sum_{j,l = 1}^n
  \frac{ \phi_{kj} } { p_j }
  \frac{ \phi_{kl} } { p_l }
  \langle \zeta_j(t) \, \zeta_l(t') \rangle
  \\
  %
  &=
  \sum_{j,l = 1}^n
  \frac{ \phi_{kj} } { p_j }
  \frac{ \phi_{kl} } { p_l }
  \left(
    p_j \, \delta_{j, l} - p_j \, p_l
  \right)
  \, \delta(t - t')
  \\
  &=
  \left(
    \sum_{j = 1}^n
    \frac{ \phi_{kj}^2 } { p_j }
    -
    \sum_{j = 1}^n \phi_{kj}
    \sum_{l = 1}^n \phi_{kl}
  \right)
  \, \delta(t - t')
  \\
  &=
  (1 - 0 \cdot 0) \, \delta(t - t').
  \end{aligned}
  $$
  where,
  we have used
  Eq. \eqref{eq:eta_def}
  on the first line,
  %
  Eq. \eqref{eq:zz_perfect}
  on the second line,
  and
  Eq. \eqref{eq:eig_orthonormal_rows}
  on the last line.
  %
  Note for $k \ge 2$,
  $$
  \sum_{j = 1}^n \phi_{kj}
  =
  \sum_{j = 1}^n \frac{ \phi_{kj} \, \phi_{1j} } { p_j }
  =
  \delta_{k1}
  =
  0.
  $$
  as the mode $1$, or the first row, $\phi_{1j}$,
  is given by the equilibrium distribution $p_j$.
}

Another special case is the one-step process\cite{vankampen}:
in each step, the system hops to the previous bin $i - 1$
or the next bin $i + 1$ with equal probability $\frac 1 2$,
ignoring out-of-boundary transitions.
%
The transition matrix is
%
\begin{equation}
\arraycolsep=3.6pt\def\arraystretch{1.4}
\mathbf T
=
\left(
  \begin{array}{cccccccc}
    \frac 1 2 & \frac 1 2 & 0 & \dots & 0 \\
    \frac 1 2 & 0         & \frac 1 2 & \dots & 0 \\
    \vdots & &  & & \vdots \\
    0 & \dots & \frac 1 2 & 0  & \frac 1 2 \\
    0 & \dots & 0 & \frac 1 2 & \frac 1 2
  \end{array}
\right).
\label{eq:T_tridiag}
\end{equation}
%
We shall further assume that
the eigenvectors of $\mathbf w$
coincide with those of $\mathbf T$,
while the eigenvalues may differ
(this is true for a class of matrices
discussed in Sec. \ref{sec:band-matrix}).
%
The eigenvalues of $\mathbf T$ are
$$
\Lambda_k = 1 - 2 \sin^2 \frac{ (k - 1) \, \pi } { 2 \, n }.
$$
Note that these $\Lambda_k$'s are
not to be confused with the eigenvalues of $\mathbf w$,
$\lambda_k$'s.
%
The correlation function $\kappa_k(t)$
decays as $\kappa_k(0) \, \Lambda_k^t$,
with the same-time value $\kappa_k(0) = 1$ being the same as
that in perfect sampling.
%
Thus,
$\Gamma_k$ is roughly twice
the autocorrelation time,
and from Eq. \eqref{eq:Gamma_sum}, we get
%
\begin{equation}
\Gamma_k
=
1 + 2 \, \sum_{t = 1}^\infty \Lambda_k^t
=
\frac{ 1 + \Lambda_k } { 1 - \Lambda_k }
=
\cot^2 \left[ \frac{ (k - 1) \, \pi } { 2 \, n } \right].
\label{eq:Gamma_onestep}
\end{equation}
%
We can use this case to model local sampling processes.

\subsubsection{\label{eq:eqlerr}
Equilibrium error under a fixed updating magnitude
}

Before addressing the optimal $\alpha(t)$,
let us first compute the equilibrium value of $\Err$,
under a fixed $\alpha(t) \equiv \alpha_0$.
%
In this case, we can drop $\Err_R$, and
%
\begin{align}
  \Err
  &=
  \Err_A
  =
  \int_0^t
    \sum_{k = 2}^n
      \Gamma_k \, (\lambda_k \, \alpha_0)^2 \,
      e^{ 2 \, \lambda_k \, \alpha_0 \, (t' - t) }
    \, dt
  \notag
  \\
  &
  \stackrel{ t \to \infty }
  { =\joinrel=\joinrel=\joinrel= }
  %
  \sum_{ k = 2 }^n
    \frac 1 2 \, \Gamma_k \, \lambda_k \, \alpha_0,
  \label{eq:error_eql}
\end{align}
%
or, in terms of the components,
%
\begin{equation}
  \left\langle
    y_k^2
  \right\rangle
  =
  \frac 1 2 \, \Gamma_k \, \lambda_k \, \alpha_0,
  \label{eq:y2_eql}
\end{equation}
%

If we can assume that before entering the asymptotic regime,
the system has entered the equilibrium under a fixed $\alpha_0$,
we can then use Eq. \eqref{eq:y2_eql}
for the $\langle y_k^2(0) \rangle$'s in Eq. \eqref{eq:error_res}.

\note{If we can interpret $\Gamma_k$ as twice the autocorrelation time,
  and the kernel is exponential
  $$
  \kappa_k = \exp\left( - \frac{2}{\Gamma_k} |t| \right),
  $$
  then a slightly more accurate value is given by\cite{vankampen}
  $$
  \left\langle
    y_k^2
  \right\rangle
  =
  \frac{      ( \lambda_k \, \alpha_0 )^2     }
       { \lambda_k \, \alpha_0 + 2 / \Gamma_k }.
  $$
  But it appears that the main error of Eq. \eqref{eq:y2_eql}
  is an underestimate instead of an overestimate.
  And a possible source of error might be from
  converting a discrete sum to a continuous integral.
}

%
%We can study two limiting cases.
%
%For $\tau \ll t$,
%the dominant contribution comes from
%the term(s) with the smallest eigenvalue $\lambda_k = \lambda_n$,
%and
%$$
%\ddot u_n(\tau) = 0,
%\quad
%\dot u_n(\tau) = c,
%\quad
%\mathrm{and}\;
%u_n(\tau) = c \, (\tau + t_0),
%$$
%which means
%$$
%\alpha(\tau) \approx \frac{1}{\lambda_n \, (\tau + t_0)}.
%$$

@@

We can find an analytical expression
of the optimal value of $\lambda$
in the limit of large $n$.
%
In this case,
we convert Eq.
\eqref{eq:error_asym_invt}
%and
%\eqref{eq:optimal_lambda_approx}
to an integral,
$\sum_i \to \frac{2 \, n}{\pi} \int dp$, and
$$
\begin{aligned}
\Err_2
&=
\frac{2 \, n}{\pi \, t}
\int_0^{\pi/2}
  \frac{    \Gamma(p) \, \left(1 - 4 \, \mu_1 \, \sin^2 p \right)^2    }
       { \lambda \, \left(2 - 8 \, \mu_1 \, \sin^2 p - \lambda \right) }
\, dp.
\end{aligned}
$$

@@

The Gaussian kernel is commonly
used in metadynamics.
%
Below we show that this
leads to a stable updating scheme,
although it can be less effective
in reducing local noises.

@@

We have shown that
without a priori information,
the single-bin scheme is asymptotically
the best among all updating scheme
in terms of convergence.
